---
title: "Maeve plays with ETH"
author: "Maeve!"
format: html
editor: visual
---
## Data Loading, packages
```{r}
library(tidyverse) 
library(sf) 
library(here) 
library(stars) 
library(tmap) 
library(viridis) 
library(readr)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(lubridate)
library(terra)
library(exactextractr)
library(readr)
library(ggplot2)
library (kableExtra)

EPR <- read.csv('EPR-2021.csv')
EPRG <- read.csv('GeoEPR-2021.csv')
divEPR <- read.csv('ED-2021.csv')
view(EPRG)
```
# Maps (and burkina)

## Senegal Coding
```{r}

epr_sen <- EPRG %>%
  mutate(statename_clean = trimws(tolower(statename))) %>%
  filter(statename_clean == "senegal")

# Convert WKT column -> sf. If your column name is exactly "the_geom", use wkt='the_geom'.
# Most GeoEPR exports are in WGS84; if yours differs, replace 4326 appropriately.
epr_sen_sf <- st_as_sf(epr_sen, wkt = "the_geom", crs = 4326)

# Fix invalid geometries just in case
if (any(!st_is_valid(epr_sen_sf))) {
  epr_sen_sf <- st_make_valid(epr_sen_sf)
}
 
cat_col <- "group"  # <-- change to your categorical field (e.g., "group_name")

tmap_mode("view")
Senegal <- tm_shape(epr_sen_sf) +
  tm_polygons(
    col     = cat_col,
    style   = "cat",            # qualitative mapping
    palette = "Set3",           # qualitative palette (RColorBrewer)
    title   = "GeoEPR Group",
    colorNA = "grey85",         # show missing explicitly
    textNA  = "Missing"
  ) +
  tm_borders(col = "grey30", lwd = 0.6) +
  tm_view(basemaps = "OpenStreetMap") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.title.size = 0.9,
    legend.text.size  = 0.75,
    title = "GeoEPR (2021): Senegal"
  )




```
## Burkina (not included)
```{r}
# Filter for Burkina Faso (using ISO3 or country name depending on variable)
burkina <- EPR %>%
  filter(statename %in% c("Burkina Faso (Upper Volta)"))

nrow(burkina)

group_counts <- burkina %>%
  transmute(group = str_squish(as.character(group))) %>%
  filter(!is.na(group), group != "") %>%
  count(group, sort = TRUE)
# Assuming the column for ethnic groups is "group" or "ethnic" (adjust if needed)
ggplot(group_counts, aes(x = reorder(group, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Ethnic Groups in Burkina Faso (EPR 2021)",
    x = "Ethnic group",
    y = "Count of entries in EPR"
  ) +
  theme_minimal(base_size = 12)

unique(EPR$statename)
```
## Mali
```{r}
epr_mali <- EPRG %>%
  mutate(statename_clean = trimws(tolower(statename))) %>%
  filter(statename_clean == "mali")

# Convert WKT column -> sf. If your column name is exactly "the_geom", use wkt='the_geom'.
# Most GeoEPR exports are in WGS84; if yours differs, replace 4326 appropriately.
epr_mali_sf <- st_as_sf(epr_mali, wkt = "the_geom", crs = 4326)

# Fix invalid geometries just in case
if (any(!st_is_valid(epr_mali_sf))) {
  epr_mali_sf <- st_make_valid(epr_mali_sf)
}
 
cat_col <- "group"  # <-- change to your categorical field (e.g., "group_name")

tmap_mode("view")
Mali <- tm_shape(epr_mali_sf) +
  tm_polygons(
    col     = cat_col,
    style   = "cat",            # qualitative mapping
    palette = "Set3",           # qualitative palette (RColorBrewer)
    title   = "GeoEPR Group",
    colorNA = "grey85",         # show missing explicitly
    textNA  = "Missing"
  ) +
  tm_borders(col = "grey30", lwd = 0.6) +
  tm_view(basemaps = "OpenStreetMap") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.title.size = 0.9,
    legend.text.size  = 0.75,
    title = "GeoEPR (2021): Mali"
  )

Mali

```
## Niger
```{r}

epr_niger <- EPRG %>%
  mutate(statename_clean = trimws(tolower(statename))) %>%
  filter(statename_clean == "niger")

# Convert WKT column -> sf. If your column name is exactly "the_geom", use wkt='the_geom'.
# Most GeoEPR exports are in WGS84; if yours differs, replace 4326 appropriately.
epr_niger_sf <- st_as_sf(epr_niger, wkt = "the_geom", crs = 4326)

# Fix invalid geometries just in case
if (any(!st_is_valid(epr_niger_sf))) {
  epr_niger_sf <- st_make_valid(epr_niger_sf)
}
 
cat_col <- "group"  # <-- change to your categorical field (e.g., "group_name")

tmap_mode("view")
Niger <- tm_shape(epr_niger_sf) +
  tm_polygons(
    col     = cat_col,
    style   = "cat",            # qualitative mapping
    palette = "Set3",           # qualitative palette (RColorBrewer)
    title   = "GeoEPR Group",
    colorNA = "grey85",         # show missing explicitly
    textNA  = "Missing"
  ) +
  tm_borders(col = "grey30", lwd = 0.6) +
  tm_view(basemaps = "OpenStreetMap") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.title.size = 0.9,
    legend.text.size  = 0.75,
    title = "GeoEPR (2021): Mali"
  )

Niger
```
## Guinea
```{r}
epr_gin <- EPRG %>%
  mutate(statename_clean = trimws(tolower(statename))) %>%
  filter(statename_clean == "guinea")

# Convert WKT column -> sf. If your column name is exactly "the_geom", use wkt='the_geom'.
# Most GeoEPR exports are in WGS84; if yours differs, replace 4326 appropriately.
epr_gin_sf <- st_as_sf(epr_gin, wkt = "the_geom", crs = 4326)

# Fix invalid geometries just in case
if (any(!st_is_valid(epr_gin_sf))) {
  epr_gin_sf <- st_make_valid(epr_gin_sf)
}
 
cat_col <- "group"  # <-- change to your categorical field (e.g., "group_name")

tmap_mode("view")
guinea <- tm_shape(epr_gin_sf) +
  tm_polygons(
    col     = cat_col,
    style   = "cat",            # qualitative mapping
    palette = "Set3",           # qualitative palette (RColorBrewer)
    title   = "GeoEPR Group",
    colorNA = "grey85",         # show missing explicitly
    textNA  = "Missing"
  ) +
  tm_borders(col = "grey30", lwd = 0.6) +
  tm_view(basemaps = "OpenStreetMap") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.title.size = 0.9,
    legend.text.size  = 0.75,
    title = "GeoEPR (2021): Mali"
  )
guinea

```

## Mauritania
```{r}
epr_maur <- EPRG %>%
  mutate(statename_clean = trimws(tolower(statename))) %>%
  filter(statename_clean == "mauritania")

# Convert WKT column -> sf. If your column name is exactly "the_geom", use wkt='the_geom'.
# Most GeoEPR exports are in WGS84; if yours differs, replace 4326 appropriately.
epr_maur_sf <- st_as_sf(epr_maur, wkt = "the_geom", crs = 4326)

# Fix invalid geometries just in case
if (any(!st_is_valid(epr_maur_sf))) {
  epr_maur_sf <- st_make_valid(epr_maur_sf)
}
 
cat_col <- "group"  # <-- change to your categorical field (e.g., "group_name")

tmap_mode("view")
mauritania <- tm_shape(epr_maur_sf) +
  tm_polygons(
    col     = cat_col,
    style   = "cat",            # qualitative mapping
    palette = "Set3",           # qualitative palette (RColorBrewer)
    title   = "GeoEPR Group",
    colorNA = "grey85",         # show missing explicitly
    textNA  = "Missing"
  ) +
  tm_borders(col = "grey30", lwd = 0.6) +
  tm_view(basemaps = "OpenStreetMap") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    legend.title.size = 0.9,
    legend.text.size  = 0.75,
    title = "GeoEPR (2021): Mali"
  )
mauritania
```

#tables 
### mali
Only in this dataset do I really see Moors being represented as an ethnic group in mali. Moors in this case appear to be used as a catch all term for arab descendent groups (white ish or something). I imagine that the divisions between the groups are too fine grained to be effectively explained in a dataset. 
```{r}

# Filter to Senegal (account for possible naming variations)
mali <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("mali"))

# Select the key columns
Mali_table <- mali %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
Mali_table

mali_latest <- mali %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  mali_latest,
  caption = "Most Recent Power Status of Ethnic Groups in Mali (EPR 2021)"
)
```
### Senegal
This one makes sense-don’t have to do a ton of work to contextualize (haven’t been there, but generally seems accurate and relevant) also could be explained by the coding parameters. 

```{r}
# Filter to Senegal (account for possible naming variations)
senegal <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("senegal"))

# Select the key columns
senegal_table <- senegal %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
senegal_table

senegal_latest <- senegal %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  senegal_latest,
  caption = "Most Recent Power Status of Ethnic Groups in Senegal (EPR 2021)"
)
```
### BF
There was too much variation in Burkina and the general power sharing relationships have been pretty stable for a while (especially during the colonial period) with a similar language background. It makes more sense to code everyone in the same group rather than split by sub-ethnic group. Only in the past 4 years have ethnic differences been politically salient. 
```{r}
# Filter to Senegal (account for possible naming variations)
bf <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("burkina faso (upper volta)"))

# Select the key columns
bf_table <- bf %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
bf_table

bf_latest <- bf %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  bf_latest,
  caption = "Most Recent Power Status of Ethnic Groups in Senegal (EPR 2021)"
)
```
### Niger
```{r}
# Filter to Senegal (account for possible naming variations)
niger <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("niger"))

# Select the key columns
niger_table <- niger %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
niger_table

niger_latest <- niger %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  niger_latest,
  caption = "Most Recent Power Status of Ethnic Groups in niger (EPR 2021)"
)
```
### Mauritania
```{r}
# Filter to Senegal (account for possible naming variations)
mauritania <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("mauritania"))

# Select the key columns
mauritania_table <- mauritania %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
mauritania_table

mauritania_latest <- mauritania %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  niger_latest,
  caption = "Most Recent Power Status of Ethnic Groups in mauritania (EPR 2021)"
)
```
### Guinea
```{r}
guinea <- EPR %>%
  mutate(statename = str_squish(as.character(statename))) %>%
  filter(str_to_lower(statename) %in% c("guinea"))

# Select the key columns
guinea_table <- guinea %>%
  select(group, status, from, to) %>%
  arrange(group, from)

# View in console
guinea_table

guinea_latest <- mauritania %>%
  group_by(group) %>%
  slice_max(to, n = 1, with_ties = FALSE) %>%
  select(group, status, to)

kable(
  guinea_latest,
  caption = "Most Recent Power Status of Ethnic Groups in Guinea (EPR 2021)"
)
```


